---
title: Lab 5 of Advanced Regression for Linguists
author: Martijn Wieling (http://www.martijnwieling.nl)
date: "Generation date: `r format(Sys.time(), '%b %d, %Y - %H:%M:%S')`"
output :
  html_document :
    toc : true
    toc_depth: 2
---

# Introduction 

In this lab session, we will look at Dutch dialect data. Make sure you have the most recent version of `R` 64 bits installed.

# Initialization

Load the required packages and download the required files

```{r setup, message=F, echo=F}
# some custom options to generate a nice html file
options(digits = 3)
options(width=90)
library(knitr)
opts_chunk$set(cache=F, comment='#', tidy = T, tidy.opts=list(blank=T, width.cutoff=90), fig.align='center', message=F, warning=T, fig.width=8, fig.height=8, dev='svg',eval=T)
```

``` {r init}
# test if the necessary packages are installed if not, install them
packages <- c("mgcv","itsadug","car")
if (length(setdiff(packages, rownames(installed.packages()))) > 0) {
  install.packages(setdiff(packages, rownames(installed.packages())))  
}

# load the packages
library(mgcv)
library(itsadug)
library(car)

# display version information
R.version.string
packageVersion('mgcv')
packageVersion('itsadug')
packageVersion('car')

# download and load the following files containing the dataset and some functions:
download.file('http://www.let.rug.nl/wieling/statscourse/lecture5/lab/functions.R', 'functions.R')
download.file('http://www.let.rug.nl/wieling/statscourse/lecture5/lab/compareML.R', 'compareML.R')
download.file('http://www.let.rug.nl/wieling/statscourse/lecture5/lab/dialectdata.rda', 'dialectdata.rda')

source('functions.R') # custom plotting functions
load('dialectdata.rda') # Dutch dialect data, subset of 48 words
```

# Data investigation

Look at the data to discover its structure. Note that predictors were standardized (z-transformed) and log-transformed (if appropriate). The dependent variable (`LD`) was centered. To z-transform a variable, you can use `scale(dat$predictor)`. To center a variable, you can use `scale(dat$predictor,scale=F)`. To log-transform you can use `log(dat$predictor)`. You can check if log-transorming is necessary by plotting the density of the predictor with `plot(density(dat$predictor))`. 

## Exercise 1: investigate the data yourself

```{r exercise1, fig.width=5, fig.height=5}
# your code goes here
```

# A first generalized additive model

We first create a simple generalized additive model in which we assess the influence of geography (i.e. the interaction between `Longitude` and `Latitude`). As `Longitude` and `Latitude` are on the same scale, we can use `s()`. 

```{r firstmodel}
model1 = bam(LD ~ s(Longitude,Latitude), data=dialectdata)
summary(model1)
```

## Visualization

To visualize the partial effect of geography (i.e. excluding all other effect in the model summary: in this case only the non-significant intercept), we can use `pvisgam`

```{r firstplot, fig.width=8,fig.height=8}
pvisgam(model1,view=c('Longitude','Latitude'),select=1,color='terrain', too.far=0.05)
```

To visualize the complete effect of geography, we can use `vis.gam` of the `mgcv` package, or `fvisgam` of the `itsadug` package (the advantage of the latter is that it allows one to exclude the random effects. Unfortunately, the CRAN version of `fvisgam` has some bugs, which prevents it from removing points which are too far (parameter `too.far`), therefore we use the `itsadug` function `fadeRug` too focus only on the points for which there is data. 

```{r secondplot, fig.width=10, fig.height=5}
par(mfrow=c(1,2))
vis.gam(model1, plot.type="contour", color="terrain", too.far=0.05, view=c("Longitude","Latitude"), main="vis.gam")
fvisgam(model1, plot.type="contour", color="terrain", too.far=0.05, view=c("Longitude","Latitude"), main="fvisgam")
fadeRug(dialectdata$Longitude,dialectdata$Latitude)
```

As an alternative one can also install the most recent version of `itsadug` in which this bug has been corrected, by executing `detach("package:itsadug", unload=T)` and `install.packages("http://www.sfs.uni-tuebingen.de/~jvanrij/packages/itsadug_1.0.7.tar.gz", type="source")`. Note that in Windows, [Rtools](https://cran.r-project.org/bin/windows/Rtools/) needs to be installed before installing packages from source.

## A note about basis functions

In the course, we have not talked about basis functions. When we use `s()` to model non-linearities, the underlying (default) basis function is a thin plate regression spline (i.e. `bs='tp'`). This is generally suitable, but there are different basis functions, which are frequently quicker to compute (e.g., a cubic spline basis: `bs='cr'`). For a tensor product spline `te()`, the default basis is `bs='cr'`. See http://www.inside-r.org/r-doc/mgcv/smooth.terms for more information. 
Also note that the default maximum complexity of the smooth differs between `te()` (where it is 5 per dimension) and `s()` (where it is about 10 per dimension).

# A three-dimensional generalized additive model
We now try to model the interaction between longitude, latitude and word frequency. 

```{r model2}
model2 <- bam(LD ~ te(Longitude,Latitude,WordFreq,d=c(2,1)), data=dialectdata, discrete=T)
summary(model2)
```

We can assess if `model1` is better than `model2` by refitting using `ML` and using `compareML`.

```{r compare1}
model1.ml <- bam(LD ~ s(Longitude,Latitude), data=dialectdata, method='ML')
model2.ml <- bam(LD ~ te(Longitude,Latitude,WordFreq,d=c(2,1)), data=dialectdata, method='ML')
compareML(model1.ml,model2.ml)
```

## Visualization

We use fvisgam to visualize the geographical pattern for low frequency words (`WordFreq = -2`), mean frequency words (`WordFreq = 0`), and high frequency words (`WordFreq = 2`).

```{r plot2, fig.width=10, fig.height=3.5}
par(mfrow=c(1,3))
fvisgam(model2, color="terrain", too.far=0.05, cond=list(WordFreq=-2), view=c("Longitude","Latitude"), main="Effect of Geography for LF words")
fadeRug(dialectdata$Longitude,dialectdata$Latitude)
fvisgam(model2, color="terrain", too.far=0.05, cond=list(WordFreq=0), view=c("Longitude","Latitude"), main="Effect of Geography for MF words")
fadeRug(dialectdata$Longitude,dialectdata$Latitude)
fvisgam(model2, color="terrain", too.far=0.05, cond=list(WordFreq=+2), view=c("Longitude","Latitude"), main="Effect of Geography for HF words")
fadeRug(dialectdata$Longitude,dialectdata$Latitude)
```

## Exercise 2: investigate the age effect

Create a model (`model2b`) to investigate the varying effect of geography only depending on population age (`PopAge`). Assess if the model is better than `model1`. Also visualize the results of the model.

```{r exercise2, fig.width=10, fig.height=3.5}
# your code goes here
```

# Creating the random-effects structure

We first add a random intercept for word to `model2`. 

```{r model3}
model3 = bam(LD ~ te(Longitude,Latitude,WordFreq,d=c(2,1)) + s(Word,bs="re"), data=dialectdata, discrete=T)
summary(model3)
```

The summary immediately shows that the random intercept for word is necessary, as it is significant.

## Visualization of a random intercept

We can visualize the random intercepts as follows.

```{r plotIntercepts, fig.width=6, fig.height=6}
plotIntercept.gam(model3,dialectdata,"Word") 
```

## Adding a random slope

We now add a fixed-effect predictor population count (`PopCnt`) and assess if it needs to be included as a by-word random slope as well.

```{r model4}
model4 = bam(LD ~ te(Longitude,Latitude,WordFreq,d=c(2,1)) + PopCnt + s(Word,bs="re") + s(Word,PopCnt,bs='re'), data=dialectdata, discrete=T)
summary(model4)
```

The summary shows that the predictor is needed as a fixed-effect predictor as well as a by-word random slope. 

## Visualization of a random slope

We can visualize the random slopes as follows.

```{r plotSlopes, fig.width=6, fig.height=6}
plotSlope.gam(model4,dialectdata,"Word","PopCnt") 
```

## Exercise 3: location as a random-effect factor

Add a random intercept for location and and visualize the results. Subsequently add a by-location random slope for WordFrequency. Is this random slope significant? Why do you think this is the case?

```{r exercise3, fig.width=6, fig.height=6}
# your code goes here
```

**Answer**: ...

# Final remarks

* Note that the (geographical) results are different than reported in the [paper](http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0023613). This is due to the relatively small dataset (containing less than 10% of the locations).
* The original dataset and all analyses can be found in the [paper package](http://openscience.uni-leipzig.de/index.php/mr2/article/view/44)
* For the Tuscan data presented during the lecture, these can be found in [another paper package](http://openscience.uni-leipzig.de/index.php/mr2/article/view/41)

# Replication
To generate the output of the analysis presented above, first install [Pandoc](http://johnmacfarlane.net/pandoc/). Then copy the following lines to the most recent version of R. 

```{r knitr, eval=F}
# install rmarkdown package if not installed
if(!"rmarkdown" %in% rownames(installed.packages())) {
   install.packages("rmarkdown")
}
library(rmarkdown) # load rmarkdown package

# download original file if not already exists (to prevent overwriting)
if (!file.exists('lab5.Rmd')) { 
  download.file('http://www.let.rug.nl/wieling/statscourse/lecture5/lab/lab5.Rmd', 'lab5.Rmd')
} 

# generate output
render('lab5.Rmd') # generates html file with results

# view output in browser
browseURL(paste('file://', file.path(getwd(),'lab5.html'), sep='')) # shows result
```