---
title: "Rnotes"
author: "Robert A. Stevens"
date: "Wednesday, January 06, 2016"
output: html_document
---

http://www.stat.columbia.edu/~gelman/bda.course/Rnotes.pdf

# Class 1: Getting started with R

You must read this document while sitting at the computer with Rstudio open. You will learn by typing in code from these chapters. You can learn further by playing with the code.

Computing is central to modern statistics at all levels, from basic to advanced. If you already know how to program, great. If not, consider this a start.

We do our computing in the open-source package R, a command-based statistical software environment that you can download and operate on your own computer, using Rstudio, a free graphical user interface for R.

**Downloading and setting up R and Rstudio**

Go to the home of R, http://www.r-project.org/ and click on the "download R" link. This will take you to a list of mirror sites. Choose any of these. Now click on the link under "Download and Install R" at the top of the page for your operating system (Linux, Mac, or Windows) and download the binaries corresponding to your system. Follow all the instructions. Default settings should be fine.

Go to the home of Rstudio, http://www.rstudio.com, click on "Download RStudio," click on "Download RStudio Desktop," and click on the installer for your platform (Windows, Mac, etc.) Then a download will occur.

When the download is done, click on the RStudio icon to start an R session.

To check that R is working, go to the Console window within RStudio. There should be a ">" prompt. Type 2 + 5 and hit Enter. You should get "[1] 7". From now on, when we say "type" _____, we mean type (or copy-and-paste) _____ into the R console and hit Enter.

You will also want to install some R packages, in particular, foreign (which allows you to read in files formatted for various other statistics packages, including Stata), knitr (which allows you to process certain R documentation), and arm (which has convenient functions for fitting and displaying some regression models). You can do this from Rstudio by clicking on the Tools tab and then on Install packages and then entering the names of the packages and clicking Install. When you want to use a package, you load it during your R session as needed, for example by typing library(foreign) or library(arm).

Try a few things, typing these one line at a time and looking at the results on the console:

```{r, comment=""}
1/3
sqrt(2)
curve(x^2 + 5, from = -2, to = 2)
```

These will return, respectively, 0.3333333, 1.414214, and a new graphics window plotting the curve y = x^2 + 5.

Finally, quit your R session by closing the Rstudio window.

**Calling functions and getting help**

Open R and play around, using the assignment function ("<-"). To start, type the following lines into your script.R file and copy-and-paste them into the R window:

```{r, comment=""}
a <- 3
print(a)
b <- 10
print(b)
a + b
a*b
exp(a)
10^a
log(b)
log10(b)
a^b
round(3.435, 0)
round(3.435, 1)
round(3.435, 2)
```

R is based on functions, which include mathematical operations (exp(), log(), sqrt(), and so forth) and lots of other routines (print(), round(), ...).

The function c() combines things together into a vector. For example, type c(4, 10, -1, 2.4) in the R console or copy-and-paste the following:

```{r, comment=""}
x <- c(4, 10, -1, 2.4)
print(x)
```

The function seq() creates an equally-spaced sequence of numbers; for example, seq(4, 54, 10) returns the sequence from 4, 14, 24, 34, 44, 54. The seq() function works with non-integers as well: try seq(0, 1, 0.1) or seq(2, -5, -0.4). For integers, a:b is shorthand for seq(a, b, 1) if b > a, or seq(b, a, -1) if b < a. Let's try a few more commands:

```{r, comment=""}
c(1, 3, 5)
1:5
c(1:5, 1, 3, 5)
c(1:5, 10:20)
seq(2, 10, 2)
seq(1, 9, 2)
```

You can get help on any function using "?" in R. For example, type ?seq. This should open an internet browser window with a help file for seq(). R help files typically have more information than you'll know what to do with, but if you scroll to the bottom of the page you'll find some examples that you can cut and paste into your console.

Whenever you are trying out a new function, we recommend using "?" to view the help file and running the examples at the bottom to see what happens.

**Sampling and random numbers**

Here's how to get a random number, uniformly distributed between 0 and 100:

```{r, comment=""}
runif(1, 0, 100)
```

And now 50 more random numbers:

```{r, comment=""}
runif(50, 0, 100)
```

Suppose we want to pick one of three colors with equal probability:

```{r, comment=""}
color <- c("blue", "red", "green")
sample(color, 1)
```

Suppose we want to sample with unequal probabilities:

```{r, comment=""}
color <- c("blue", "red", "green")
p <- c(0.5, 0.3, 0.2)
sample(color, 1, prob = p)
```

Or we can do it all in one line, which is more compact but less readable:

```{r, comment=""}
sample(c("blue", "red", "green"), 1, prob = c(0.5, 0.3, 0.2))
```

**Data types**

*Numeric data*. In R, numbers are stored as numeric data. This includes many of the examples above as well as special constants such as pi.

*Big and small numbers*. R recognizes scientific notation. A million can be typed in as 1000000 or 1e6, but not as 1,000,000. (R is particular about certain things. Capitalization matters, “," doesn't belong in numbers, and spaces usually aren't important.) Scientific notation also works for small numbers: 1e-6 is 0.000001 and 4.3e-6 is 0.0000043.

*Infinity*. Type (or copy-and-paste) these into R, one line at a time, and see what happens:

```{r, comment=""}
1/0
-1/0
exp(1000)
exp(-1000)
1/Inf
Inf + Inf
-Inf - Inf
0/0
Inf - Inf
```

Those last two operations return NaN (Not a Number); type ?Inf for more on the topic. In general we try to avoid working with infinity but it is convenient to have Inf for those times when we accidentally divide by 0 or perform some other illegal mathematical operation.

*Missing data*. In R, NA is a special keyword that represents missing data. For more information, type ?NA. Try these commands in R:

```{r, comment=""}
NA
2 + NA
NA - NA
NA/NA
NA*NA
c(NA, NA, NA)
c(1, NA, 3)
10*c(1, NA, 3)
NA/0
NA + Inf
is.na(NA)
is.na(c(1, NA, 3))
```

The is.na() function tests whether or not the argument is NA. The last line operates on each element of the vector and returns a vector with three values that indicate whether the corresponding input is NA.

*Character strings*. Let's sample a random color and a random number and put them together:

```{r, comment=""}
color <- sample(c("blue", "red", "green"), 1, prob = c(0.5, 0.3, 0.2))
number <- runif(1, 0, 100)
paste(color, number)
```

Here's something prettier:

```{r, comment=""}
paste(color, round(number,0))
```

**TRUE, FALSE, and ifelse**

Try typing these:

```{r, comment=""}
2 + 3 == 4
2 + 3 == 5
1 < 2
2 < 1
```

In R, the expressions ==, <, > are comparisons and return a logical value, TRUE or FALSE as appropriate. Other comparisons include <= (less than or equal), >= (greater than or equal), and != (not equal).

Comparisons can be used in combination with the ifelse() function. The first argument takes a logical statement, the second argument is an expression to be evaluated if the statement is true, and the third argument is evaluated if the statement is false. Suppose we want to pick a random number between 0 and 100 and then choose the color red if the number is below 30 or blue otherwise:

```{r, comment=""}
number <- runif(1, 0, 100)
color <- ifelse(number < 30, "red", "blue")
```

**Loops**

A key aspect of computer programming is looping - that is, setting up a series of commands to be performed over and over. Start by trying out the simplest possible loop:

```{r, comment=""}
for (i in 1:10){
  print("hello")
}
```

Or:

```{r, comment=""}
for (i in 1:10){
  print(i)
}
```

Or:

```{r, comment=""}
for (i in 1:10){
  print(paste("hello", i))
}
```

The curly braces define what is repeated in the loop.

Here's a loop of random colors:

```{r, comment=""}
for (i in 1:10){
  number <- runif(1, 0, 100)
  color <- ifelse(number < 30, "red", "blue")
  print(color)
}
```

# Class 2: Reading, writing, and looking at data in R

**Your working directory**

Choose a working directory on your computer where you will do your R work. Suppose your working directory is c:/myfiles/stat/. Then you should put all your data files in this directory, and all the files and graphs you save in R will appear here too. To set your working directory in Rstudio, click on the Session tab and then on Set Working Directory and then on Choose Directory and then navigate from there.

Rstudio has several subwindows: a text editor, the R console, a graphics window, and a help window. It's generally best to type your commands into the Rstudio text editor and then copy and paste into the R console. When your session is over, you can save the contents of the text editor into a plain-text file with a name such as todays_work.R that you can save in your working directory.

Now go to the R console and type getwd(). This shows your current R working directory. Change your working directory by typing setwd("c:/myfiles/stat/"), or whatever you would like to use.

Then type getwd() in the R console. This should return your working directory (for example, c:/myfiles/stat/).

**Reading data**

Let's read some data into R. The file heads.csv has data from a coin flipping experiment done in a previous class. Each student flipped a coin 10 times and we have a count of the number of students who saw exactly 0, 1, 2, ..., 10 heads. The file is located at the website http://www.stat.columbia.edu/~gelman/bda.course/. Start by going to this location, downloading the file, and saving it as heads.csv in your working directory (for example, c:/myfiles/stat/).

Now read the file into R:

```{r, comment=""}
setwd("~/_comm")
heads <- read.csv("heads.csv")
```

R can also read the file directly from the web rather than saving it to your computer first. For example:

```{r, comment=""}
heads <- read.csv("http://www.stat.columbia.edu/~gelman/bda.course/heads.csv")
```

Typing the name of any object in R displays the object itself. So type heads and look at what comes out.

Now for a longer file: download nycData.csv from http://www.stat.columbia.edu/~gelman/bda.course/, save it in your working directory, and read it in:

```{r, comment=""}
nyc <- read.csv("nycData.csv")
```

This data is a subset of U.S. census data from New York City zip codes. If you now type nyc from the console, the file will scroll off the screen: it's too big! Instead, let's just look at the first 5 rows:

```{r, comment=""}
nyc[1:5, ]
```

We'll talk in a bit about what to do with these data, but first we'll consider some other issues involving data input.

What if you have tabular data separated by spaces and tabs, rather than columns? For example, mile.txt from http://www.stat.columbia.edu/~gelman/bda.course/. You just use the read.table() function:

```{r, comment=""}
mile <- read.table("mile.txt", header = TRUE)
mile[1:5, ]
```

The header = TRUE argument is appropriate here because the first line of the file mile.txt is a "header," that is, a list of column names. If the file had just data with no header, we would simply call read.table("mile.txt") with no header argument.

**Writing data**

You can save data into a file using write instead of read. For example, to write the R object heads into a comma-separated file output1.csv, we would type write.csv(heads, "output1.csv"). To write it into a space-separated file output2.txt, it's just write.table(heads, "output2.txt").

**Examining data**

At this point, we should have three variables in our R environment, heads, nyc, and miles. We can see what we have in our session by clicking on the Environment tab in the Rstudio window.

*Data frames, vectors, and subscripting*. Most of the functions used to read data return a data structure called a data frame. You can see this by typing class(heads). Each column of a data frame is a vector. We can access the first column of heads by typing heads[,1]. Data frames are indexed using two vectors inside "[" and "]"; the two vectors are separated by a comma. The first vector indicates which rows you are interested in and the second vector indicates what columns you are interested in. For example, heads[6, 1] shows the number of heads observed and heads[6, 2] shows the number of students that observed that number of heads. Leaving it blank is shorthand for including all. Try:

```{r, comment=""}
heads[6, ]
heads[1:3, ]
heads[ ,1]
heads[ ,1:2]
heads[ , ]
```

To find the number of columns in a data frame, use the function length(). To find the number of rows, use nrow(). We can also find the names of the columns by using names(). Try these:

```{r, comment=""}
length(heads)
nrow(heads)
names(heads)
length(nyc)
nrow(nyc)
names(nyc)
```

We can also index a column within a data frame by using \$. The call to names(nyc) showed the names of the 7 columns in the data set. We can access the zip codes by typing nyc\$zipcode or the population of each of the zip codes by typing nyc\$pop.total. As we mentioned before, these columns are vectors. We can index these vectors like any other vectors. Try:

```{r, comment=""}
nyc$zipcode[1:5]
nyc$pop.total[1:5]
```

# Class 3: Making graphs in R

**Graphing data**

**Scatterplots**. A scatterplot shows the relation between two variables. The data frame called mile contains the world record times in the mile run since 1900 as three columns, year, min, and sec. For convenience we create the following derived quantities in R:

```{r, comment=""}
year <- mile$year
record <- mile$min + mile$sec/60
```

Figure 3.1 plots the world record time (in minutes) against time. We can create scatterplots in R using plot. Here is the basic graph:

```{r, comment=""}
plot(year, record, main = "World record times in the mile run")
```

**Figure 3.1a World record times (in minutes) in the mile run since 1900. The plot was made with the basic function call, plot(year, record, main = "World record times in the mile run").**

And here is a slightly prettier version:

```{r, comment=""}
par(mar = c(3, 3, 3, 1), mgp = c(2, 0.5, 0), tck = -0.01)
plot(year, record, main = "World record times in the mile run", bty = "l")
```

**Figure 3.1b World record times (in minutes) in the mile run since 1900. The plot was made with more formatting: par(mar = c(3, 3, 3, 1), mgp = c(2, 0.5, 0), tck = -0.01); plot(year, record, main = "World record times in the mile run", bty = "l").**

The par function sets graphical parameters, in this case reducing the blank border around the graph, placing the labels closer to the axes, and reducing the size of the tick marks, compared to the default settings in R. In addition, in the call to plot, we have set the box type to l, which makes an "L-shaped" box for the graph rather than fully enclosing it.

We want to focus on the basics, so for the rest of this chapter we will show simple plot calls that won't make such pretty graphs. But we though it would be helpful to show one example of a pretty graph, hence the code just shown above.

*Fitting a line to data*. We can fit a regression line predicting world record time from the date (parameterized as years since 1900) as follows:

```{r, comment=""}
years_since_1900 <- year - 1900
fit <- lm(record ~ years_since_1900)
```

We then load in the arm package so that we can display the fit. Here is the result:

```{r, comment=""}
library(arm)
display(fit)
```

You'll learn later how to interpret all these results. All that you need to know now is that the estimated coefficients are 4.33 and -0.01; that is, the fitted regression line is y = 4.33 -– 0.01(year - 1900). It will help to have more significant digits on this slope, so we type display(fit, digits = 5) to get this:

```{r, comment=""}
display(fit, digits = 5)
```

The best-fit line is y = 4.33 - 0.00655(year - 1900).

We can add the straight line to the scatterplot by adding the following line after the call to plot:

```{r, comment=""}
plot(year, record, main = "World record times in the mile run", bty = "l")
curve(4.33 - 0.00655*(x - 1900), add = TRUE)
```

**Figure 3.2a Fitted line predicting world record in the mile run given year (data shown in Figure 3.1), plotted using the code, curve(4.33 - 0.00655\*(x - 1900), add = TRUE).**

The first argument is the equation of the line as a function of x. The second argument, add = TRUE, tells R to draw the line onto the existing scatterplot. And we can graph this line by itself using the following R code:

```{r, comment=""}
curve(4.33 - 0.00655*(x - 1900), from = min(year), to = max(year)) 
```

**Figure 3.2b Fitted line predicting world record in the mile run given year (data shown in Figure 3.1), plotted using the code, curve(4.33 - 0.00655\*(x - 1900), from = min(year), to = max(year)).**

The from and to arguments specify the range of values of x over which the curve is plotted. The result is displayed in Figure 3.2a,

Multiple graphs on a page

Visualizations can be much more powerful using small multiples: repeated graphs on a similar theme. There are various ways to put multiple graphs on a page in R; one way uses the par function with its mfrow option, which tells R to lay out graphs in a grid, row by row.

We illustrate in Figure 3.3 with a simple example plotting random numbers:

```{r, comment=""}
par(mfrow = c(4, 5), mar = c(2, 1, 2, 1), pty = "s", cex = 0.5)
for (i in 1:4) {
  for (j in 1:5) {
    x <- rnorm(10)
    y <- rnorm(10)
    plot(x, y, xlim = c(-4, 4), ylim = c(-4, 4), xlab = "", ylab = "", 
         main = paste("Row", i, "Column", j))
  }
}
```

**Figure 3.3 Example of a grid of graphs of random numbers produced in R. Each graph plots 10 pairs of numbers randomly sampled from the 2-dimensional normal distribution, and the display has been formatted to make 20 plots visible on a single page with common scales on the x and y-axes and customized plot titles.**

The actual code we used to make the graphs is more complicated because we included code to label the graphs, put them on a common scale, and size them to better fit on the page, but the above code gives the basic idea.

Figure 3.4 displays a fancier version of the same principle. In this case the R code is far too involved to copy right here, but it's good to see what the system is capable of.

Figure 3.4 Example of a grid of graphs produced in R: estimates of public support for federal funding for school vouchers, as a function of income, race/ethnicity/religion, and state.

# Class 4: Working with messy data in R

**Reading in survey data, one question at a time**

Data on the heights, weights, and incomes of a random sample of Americans are available from the Work, Family, and Well-Being survey conducted by Catherine Ross in 1990. We downloaded the data file, 06666-0001-Data.txt, and the codebook 06666-0001-Codebook.txt from the Inter-university Consortium for Political and Social Research. Information on the survey is at http://dx.doi.org/10.3886/ICPSR06666 and can be downloaded if you create an account with the ICPSR.

We saved the files under the names wfw90.dat and wfwcodebook.txt in the directory http://www.stat.columbia.edu/~gelman/bda.course/.

Figure 4.1 shows the first ten lines of the data, and Figure 4.2 shows the relevant portion of the codebook. Our first step is to save the data file wfwcodebook.txt in our working directory. We then want to extract the responses to the questions of interest. To do this we first create a simple function, to read columns of data, making use of R's function read.fwf (read in fixed width format).

```{r, comment=""}
wfw90 <- read.fwf("wfw90.dat", widths = 75)
head(wfw90, 10)
```

**Figure 4.1 First ten lines of the file wfw90.dat, which has data from the Work, Family, and Well-Being survey.**

```{r, comment=""}
wfwcodebook <- read.fwf("wfwcodebook.txt", widths = 100, as.is = TRUE)
wfwcodebook$V1[736:737]
wfwcodebook$V1[787:788]
wfwcodebook$V1[791]
wfwcodebook$V1[1604:1610]
wfwcodebook$V1[1824:1829]
```

**Figure 4.2 Selected rows of the file wfwcodebook.txt, which first identifies the columns in the data corresponding to each survey question and then gives the question wordings.**

Copy the following into the R console and then you will be able to use our function for reading one variable at a time from the survey. The following code is a bit tricky so you're not expected to understand it; you can just copy it in and use it.

```{r, comment=""}
read.columns <- function (filename, columns) {
  start <- min(columns)
  length <- max(columns) - start + 1
  if (start == 1)
    return(read.fwf(filename, widths = length))
  else
    return(read.fwf(filename, widths = c(start - 1, length))[ ,2])
}
```

Now we extract the data that we are interested in.

```{r, comment=""}
height_feet <- read.columns("wfw90.dat", 144)
height_inches <- read.columns("wfw90.dat", 145:146)
weight <- read.columns("wfw90.dat", 147:149)
income_exact <- read.columns("wfw90.dat", 203:208)
income_approx <- read.columns("wfw90.dat", 209:210)
sex <- read.columns("wfw90.dat", 219)
```

The data did not come in a convenient comma-separated or tab-separated format, so we used the function read.columns to read the coded responses, one question at a time.

**Cleaning data within R**

We now must put the data together in a useful form, doing the following for each variable of interest:

1. Look at the data
2. Identify errors or missing data
3. Transform or combine raw data into summaries of interest.

We start with height, typing: table(height_feet, height_inches). Here is the result:

```{r, comment=""}
table(height_feet, height_inches)
```

Most of the data look fine, but there are some people with 9 feet and 98 or 99 inches (missing data codes) and one person who is 7 feet 7 inches tall (probably a data error). We recode these problem cases as missing:

```{r, comment=""}
height_inches[height_inches > 11] <- NA
height_feet[height_feet >= 7] <- NA
```

And then we define a combined height variable:

```{r, comment=""}
height <- 12*height_feet + height_inches
```

We do the same thing for sex, which simply yields:

```{r, comment=""}
table(sex)
```

No problems. But we prefer to have a more descriptive name, so we define a new variable, female:

```{r, comment=""}
female <- sex - 1
```

This indicator variable equals 0 for men and one for women.

Next, we type table(weight) and get the following:

```{r, comment=""}
table(weight)
```

Everything looks fine until the end. 998 and 999 must be missing data, which we duly code as such:

```{r, comment=""}
weight[weight > 500] <- NA
```

Coding the income responses is more complicated. The variable income\_exact contains exact responses for income (in dollars per year) for those who answered the question. Typing table(is.na(income\_exact)) reveals that 1380 people answered the question (that is, is.na(income\_exact) is FALSE for these respondents), and 651 did not answer (is.na was TRUE). These nonrespondents were asked the second, discrete income question, income\_approx, which gives incomes in round numbers (in thousands of dollars per year) for people who were willing to answer in this way. A careful look at the codebook reveals that an income\_approx code of 1 corresponds to people who did not supply an exact income value but did say it was more than $100,000.We code these people as having incomes of 150,000. (This is approximately the average income of the over-100,000 group from the exact income data, as calculated by mean(income\_exact[income\_exact > 100000], na.rm = TRUE), which yields the value 166,600.) The income_approx codes also appears to have several values indicating ambiguity or missingness, which we code as NA.

We create a combined income variable as follows:

```{r, comment=""}
income_approx[income_approx >= 90] <- NA
income_approx[income_approx == 1] <- 150
income <- ifelse(is.na(income_exact), 1000*income_approx, income_exact)
```

The new income variable still has 237 missing values (out of 2031 respondents in total) and is imperfect in various ways, but we have to make some choices when working with real data.

**Looking at the data**

If you stare at the table of responses to the weight question you can see more. People usually round their weight to the nearest 5 or 10 pounds, and so we see a lot of weights reported as 100, 105, 110, and so forth, but not so many in between. Beyond this, people appear to like round numbers: 57 people report weights of 200 pounds, compared to only 46 and 36 people reporting 190 and 210, respectively.

Similarly, if we go back to reported heights we see some evidence that the reported numbers do not correspond exactly to physical heights: 129 people report heights of exactly 6 feet, compared to 90 people at 5 feet 11 inches and 59 people at 6 feet 1 inch. Who are these people? Let's look at the breakdown of height and sex. Here's the result:

```{r, comment=""}
table(female, height)
```

The extra 6-footers (72 inches tall) are just about all men. But there appear to be too many women of exactly 5 feet tall and an excess of both men and women who report being exactly 5 feet 6 inches.

## Class 5: Some programming in R

**Working with vectors**

In R, a vector is a list of items. These items can include numerics, characters, or logicals. A single value is actually represented as a vector with one element. Here are some vectors:

- (1, 2, 3, 4, 5)

- (3, 4, 1, 1, 1)

- ("A", "B", "C")

Here's the R code to create these:

```{r, comment=""}
x <- 1:5
y <- c(3, 4, 1, 1, 1)
z <- c("A", "B", "C")
```

And here's a random vector of 5 random numbers between 0 and 100:

```{r, comment=""}
u <- runif(5, 0, 100)
```

Mathematical operations on vectors are done component-wise. Take a look:

```{r, comment=""}
x
y
x + y
1000*x + u
```

There are scalar operations on vectors:

```{r, comment=""}
1 + x
2*x
x/3
x^4
```

We can summarize vectors in various ways, including the sum and the average (called the "mean" in statistics jargon):

```{r, comment=""}
sum(x)
mean(x)
```

We can also compute weighted averages if we know the weights. We illustrate with a vector of 3 elements:

```{r, comment=""}
x <- c(100, 200, 600)
w1 <- c(1/3, 1/3, 1/3)
w2 <- c(0.5, 0.2, 0.3)
```

In the above code, the vector of weights w1 has the effect of counting each of the three items equally; vector w2 counts the first item more. Here are the weighted averages:

```{r, comment=""}
sum(w1*x)
sum(w2*x)
```

Or suppose we want to weight in proportion to population:

```{r, comment=""}
N <- c(310e6, 112e6, 34e6)
sum(N*x)/sum(N)
```

Or, equivalently,

```{r, comment=""}
N <- c(310e6, 112e6, 34e6)
w <- N/sum(N)
sum(w*x)
```

The cumsum() function does the cumulative sum. Try this:

```{r, comment=""}
a <- c(1, 1, 1, 1, 1)
cumsum(a)
a <- c(2, 4, 6, 8, 10)
cumsum(a)
```

**Subscripting**

Vectors can be indexed by using brackets, "“[ ]". Within the brackets we can put in a vector of elements we are interested in either as a vector of numbers or a logical vector. When using a vector of numbers, the vector can be arbitrary length, but when indexing using a logical vector, the length of the vector must match the length of the vector you are indexing. Try these:

```{r, comment=""}
a <- c("A", "B", "C", "D", "E", "F", "G", "H", "I", "J")
a[1]
a[2]
a[4:6]
a[c(1, 3, 5)]
a[c(8, 1:3, 2)]
a[c(FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, FALSE, FALSE, FALSE, FALSE)]
```

As we have seen in some of the previous examples, we can perform mathematical operations on vectors. These vectors have to be the same length, however. If the vectors are not the same length, we can subset the vectors so they are compatible. Try these:

```{r, comment=""}
x <- c(1, 1, 1, 2, 2)
y <- c(2, 4, 6)
x[1:3] + y
x[3:5]*y
y[3]^x[4]
x + y
```

The last line runs but produces a warning. These warnings should not be ignored since it isn't guaranteed that R would carry out the operation as you intended.

**Writing your own functions**

You can write your own function. Most of the functions we will be writing will take in one or more vectors and return a vector. Below is an example of a simple function that triples the value provided:

```{r, comment=""}
triple <- function(x) {
  return(3*x)
}
```

This function has one argument, x. The body of the function is within the curly braces and the arguments of the function are available for use within the braces. In our example function, we multiply x by 3 and we return it back to the user. For example, to call this function, type triple(c(1, 2, 3, 5)):

```{r, comment=""}
triple(c(1, 2, 3, 5))
```

If we wanted to have more than one argument, we could do something like this:

```{r, comment=""}
new_function <- function(x, y, a, b) {
  return(a*x + b*y)
}
new_function(1, 2, 3, 4)
```

**Optimization**

```{r, comment=""}
par(mfrow = c(1, 1), mar = c(5, 4, 4, 2) + 0.1, pty = "m", cex = 1) # reset defaults
curve(15 + 10*x - 2*x^2, from = -2, to = 5)
```

**Figure 5.1 The parabola y = 15 + 10x - 2x^2, plotted in R using curve(15 + 10\*x - 2*x^2, from = -2, to = 5). The maximum is at x = 2.5.**

*Finding the peak of a parabola*. Figure 5.1 shows the parabola y = 15 + 10x - 2x^2. As you can see, the peak is at x = 2.5. How can we find this solution systematically (and without using calculus, which would be cheating)? Finding the maximum of a function is called an optimization problem. Here's how we do it in R.

1. Graph the function, in this case, curve(15 + 10\*x - 2\*x^2, from = x1, to = x2), where x1 and x2 are numbers. Play around with the "from" and "to" arguments until the maximum appears.

```{r, comment=""}
curve(15 + 10*x - 2*x^2, from = 2, to = 3)
```

2. Write it as a function in R:

```{r, comment=""}
parabola <- function(x) {
  return(15 + 10*x - 2*x^2)
}
```

3. Call optimize(). We can find the maximum of a function in R through the optimize() function which takes the function to optimize as the first argument, the interval to optimize over as the second argument, and an optional argument indicating whether or not you are searching for the function's maximum. For the example above, suppose we are interested in maximizing over the range from x = -100 to x = 100. Then we can call:

```{r, comment=""}
optimize(parabola, interval = c(-100, 100), maximum = TRUE)
```

This returns two values, x and f(x). The value labeled maximum is the x at which the function is optimized and the objective is its corresponding f(x) value.

4. Check the solution on the graph to see that it makes sense.

*Restaurant pricing*. Suppose you own a small restaurant and are trying to decide how much to charge for dinner. For simplicity, suppose that dinner will have a single price, and that your marginal cost per dinner is $11. From a marketing survey, you have estimated that, if you charge $x for dinner, the average number of customers per night you will get is 5000 = x^2. How much should you charge, if your goal is to maximize expected net profit?

Your expected net profit is the expected number of customers times net profit per customer; that is, (5000/x^2)*(x - 11). To get a sense of where this is maximized, we can first make a graph:

```{r, comment=""}
net_profit <- function(x){
  return((5000/x^2)*(x - 11))
}
curve(net_profit(x), from = 10, to = 100, xlab = "Price of dinner", ylab = "Net profit per night")
```

From a visual inspection of this curve, the peak appears to be around x = 20. There are two ways we can more precisely determine where net profit is maximized.

First, the brute-force approach. The optimum is clearly somewhere between 10 and 100, so let's just compute the net profit at a grid of points in this range:

```{r, comment=""}
x <- seq(10, 100, 0.1)
y <- net_profit(x)
max(y)
```

The maximum value of net profit is then simply max(y), which equals 113.6, and we can find the value of x where profit is maximized:

```{r, comment=""}
x[y == max(y)]
```

It turns out to be 22. (The value 22 is x[121], that is, the 122th element of x.) Above, we are subscripting the vector x with a logical vector y == max(y), which is a vector of 900 FALSE's and one TRUE at the maximum.

We can also use the optimize() function:

```{r, comment=""}
optimize(net_profit, interval = c(10, 100), maximum = TRUE)
```

The argument interval gives the range over which the optimization is computed, and we set maximum = TRUE to maximize (rather than minimize) the net profit.
