---
title: "Sets"
author: "Robert A. Stevens"
date: "August 3, 2016"
output: html_document
---

https://stat.ethz.ch/R-manual/R-devel/library/base/html/sets.html

## Set Operations

Examples

```{r, comment=NA}
(x <- c(sort(sample(1:20, 9)), NA))
(y <- c(sort(sample(3:23, 7)), NA))
union(x, y)
intersect(x, y)
setdiff(x, y)
setdiff(y, x)
setequal(x, y)

## True for all possible x & y :
setequal(union(x, y), c(setdiff(x, y), intersect(x, y), setdiff(y, x)))

is.element(x, y) # length 10
is.element(y, x) # length  8
```

https://www.safaribooksonline.com/library/view/the-art-of/9781593273842/ch08s05.html

## Set Operations

R includes some handy set operations, including these:

- union(x, y): Union of the sets x and y

- intersect(x, y): Intersection of the sets x and y

- setdiff(x, y): Set difference between x and y, consisting of all elements of x that are not in y

- setequal(x, y): Test for equality between x and y

- c %in% y: Membership, testing whether c is an element of the set y

- choose(n, k): Number of possible subsets of size k chosen from a set of size n

Here are some simple examples of using these functions:

```{r, comment=NA}
x <- c(1, 2, 5)
y <- c(5, 1, 8, 9)
union(x, y)
intersect(x, y)
setdiff(x, y)
setdiff(y, x)
setequal(x, y)
setequal(x, c(1, 2, 5))
2 %in% x
2 %in% y
choose(5, 2)
```
